Gap formation and stability in non-isothermal protoplanetary discs

### 2015-03-27 paper/results.tex

\subsection{Gap structure}
We first compare the gap structures formed by planet-disc
interaction as a function of the cooling time. The azimuthally-averaged 
gap profiles are shown in Fig. \ref{intial1D} for different values of 
$\tilde\beta$. Gaps formed with lower $\tilde\beta$ (faster cooling) 
are deeper with steeper gradients at the gap edges. Faster cooling rates also 
increase (decrease) the surface density maxima (minima). However, a
clean gap does not form in this short time period.%

@@

@@

However, we remark that the change in the gap structure becomes less 
significant at long cooling times, as Fig. \ref{intial1D} shows that
the profiles with $\tilde{\beta}=1$ and $\tilde{\beta}=10$ are 
similar. This implies that the effect of cooling timescale on the
RWI through the set up of the gap profile, becomes less important for
large $\tilde{\beta}$.

@@

\subsection{Axisymmetric stability}
The initial planet-disc interaction form bumps and grooves in the gap profiles
which can potentially be unstable due to axisymmetric instabilities. The
generalised local axisymmetric stability condition is the Solberg-Hoiland
criterion,  
\begin{align}
  \kappa^2+N^2 \geq 0 
\end{align}
where
\begin{align}
 N^2=\frac{1}{\Sigma} \frac{\partial P}{\partial r}
 \left(\frac{1}{\Sigma} \frac{\partial \Sigma}{\partial
     r}-\frac{1}{\gamma P} \frac{\partial P}{\partial r}  \right) 
\end{align}
is the square of the Brunt-V\"ais\"al\"a frequency. 
At the outer gap edge $r=r_p+2.5r_h$,  where the RWI is excited
(see below),  we find $\kappa^2 + N^2$ reaches local minimum with a value
$\sim 0.45 \, \Omega^2(r_p)$
%$\sim 2\times10^{-4}$ (code units)
 for all $\tilde\beta$. The
Brunt-V\"ais\"al\"a frequency at the outer gap edge is
$N\sim 0.1 \, \Omega(r_p)$,  
% $N\sim 10^{-5}$,
decreasing marginally with longer cooling rate. 
The Solberg-Hoiland criteria is similarly satisfied for the entire 2D disc throughout the 
simulations.
Thus for all values of $\tilde\beta$ the planet-induced gaps are
stable to axisymmertic {\bf perturbations}.

\subsection{Non-axisymmetric instability}\label{linear}
We now examine the evolution of the gap for $t>30P_0$, with the
planet potential switched off, but with an added surface density
perturbation. For all three cooling times $\tilde{\beta}=0.1,\, 1.0,\,
10$, we observe exponential growth of non-axisymmetric
structures. An example is shown in Fig. \ref{linearmodes} for 
$\tilde{\beta}=10$. We characterize these
modes with an azimuthal wavenumber $m$ and growth rate $\sigma(m)$ as defined by
Eq.~\ref{fouriertransform}---\ref{growth}. Mode amplitudes were
averaged over $r-r_p\in[2,5]r_h$. Table \ref{modetable}
lists the growth rates measured during 
linear growth for 5 values of $m$ centred around that with maximum
growth rate.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=1.2cm
  0cm 0cm 0cm]{figures/linear_stability}
  \caption{Evolution of azimuthal Fourier modes of disc surface
    density, non-dimensioqnalized by the initial axisymmetric component 
    $\Sigma_0(t=0)$, for the
    `planet-off' simulation with $\tilde{\beta}=10$. Colours correspond
    to different $m$ values. The $m=3$ component is the fastest growing
    mode during linear growth with a growth rate of 
    $\gamma=0.017\Omega(r_p)$.\label{linearmodes}}
  % {\bf is the $\Sigma_0$ at t=0?} }  
\end{figure}

@@

\subsection{Long term evolution} \label{nonlinearplanetoff}

@@

This decay is probably due to numerical viscosity. During the
  slow decay the vortex elongates (weakens) while its radial width
  remains $O(H)$, so its surface density decreases. In addition, for 
  $\tilde{\beta}=0.1,\,0.5$ and $1.0$ we also observe the appearance of
  spiral waves associated with the vortex, which may contribute to
  its dissipation (see below). 
We will see in the
next section that this decay after linear growth is very
different to when the planet potential is kept on.

@@

\section{Non-linear evolution of
  gap-edge vortices with finite cooling time}\label{nonlinear} 
We now examine long-term simulations of gap-edge vortices for
$\tilde{\beta}=0.1,0.5,1,5,10$. (Additional cases are presented in 
  \S\ref{lifetime_discuss} when examining vortex lifetimes as a
  function of $\tilde{\beta}$.)  The planet potential is kept on
throughout. We employ a grid with $(N_r,N_{\phi})=(512,1024)$ in order
for these 
simulations to be computationally feasible. We also use a larger
disc with $r_{\mathrm{out}}=45r_{\mathrm{in}}$ to minimise boundary
effects on vortex evolution, and apply open boundaries at
$r=r_\mathrm{in},\,r_\mathrm{out}$.

We comment that lower-resolution simulations with
$(N_r,N_{\phi})=(256,512)$ show similar behaviour and trends as the
high-resolution runs reported below. % {\bf These simulations
  % include a comparison between a run with $\tilde{\beta}=0.01$ with a 
  % locally isothermal disc, which show similar behavior. In particular,
  % we find vortex lifetimes have a non-monotonic dependence on
  % $\tilde{\beta}$, with longest lifetimes as $\tilde{\beta}\to 0$ (or
  % locally isothermal discs) and for $\tilde{\beta} = O(1)$}.

@@

In all cases the system remains in a quasi-steady state for
$\gtrsim800P_0$ with a single vortex circulating 
the outer gap edge at the local Keplerian  
frequency. 
Fig. \ref{Vortex2D} shows a typical plot of the relative 
surface density perturbation in this state. During this stage, the 
vortex intensifies. This is better shown in Fig. \ref{rossbyplot} as the
evolution of Rossby numbers measured at the vortex centres.  
The Rossby number increases in magnitude during quasi-steady state,
but the maximum $|Ro|$ is similar for all $\tilde{\beta}$: the vortex
reaches a characteristic value of $Ro\approx-0.35$ for 
$\tilde\beta=0.1$ and $Ro\approx -0.45$ for $\tilde\beta=10$.

@@

\begin{figure}
  \includegraphics[width=\linewidth,height=\linewidth]{figures/vortex2D}
  \caption{Relative surface density perturbation for the
    $\tilde\beta=0.1$ case during quasi-steady state with a single
    vortex at the outer gap edge. The plot for other values of the cooling time
    $\tilde{\beta}$ are similar. The decrease in the surface
      density near $r\sim 40 r_\mathrm{in}$ arises from mass loss due to the open
      boundary condition imposed at $r_\mathrm{out}$.
    \label{Vortex2D} }
\end{figure}

@@

@@

@@

In quasi-steady ($t=700P_0$) the vortex is elongated with a vortex aspect ratio $ \approx 4$
, but becomes more compact approaching a ratio of 2 during its decay ($t=1310P_0$).   
Notice in Fig. \ref{shockplot} the appearance of wakes extending from
either side of the vortex at $t=1300P_0$. These 
wakes correlate with large gradients in surface density (bottom
panel), and are first seen in the later half of the quasi-steady
state. We find the time at which 
the vortex begins to decay coincides with the emergence of these
wakes.

During the quasi-steady state the vortex 
  orbits at $r\sim r_p+6r_h$. We do not see significant
  vortex migration at this stage, since the vortex is located at a 
  surface density maximum \citep{paardekooper10}. However,
  simultaneous with the appearance of the wakes, we observe the vortex
  begins to migrate inwards to $r\sim r_p+5r_h$.

% As the vortices grow in intensity wakes are seen to extend from them. 
During quasi-steady state the average value of the surface density gradient
along the wakes is $|w_s\nabla\Sigma/\Sigma| \sim 0.4 $, where 
$w_s\simeq0.1$ (code units) is a typical length scale of 
the surface density variation across the wake.   
Just before the $m=1$ amplitude begins to decrease, we observe this quantity
sharply increases to $ \sim 0.6 $, and remains around this value
until the vortex dies out, at which point the associated Rossby number
begins decreasing to zero. After the 
  vortex reaches small amplitudes ($1\lesssim\Delta\Sigma/\Sigma$), it 
  migrates out to $r\sim r_p+ 6.5 r_h$.

%{\bf :check}.

@@

\begin{figure}
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/shock1}
  }
  \hfill
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/shock2}
  }
  \hfill
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/shock3}
  } \\[-0.98cm]
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/shock4}
  }
  \hfill
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/shock5}
  }
  \hfill
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/shock6}
  } 
  \caption{The vortex in the case with $\tilde{\beta}=1$
    during quasi-steady state (left), start of decay 
    (middle), and just after the decay in the $m=1$ amplitude
    (right). The surface density perturbation
    (top) and the associated surface density gradient (bottom) are
    shown. Wake-like features corresponding to large density gradients
    are found to originate from the vortices during the late phase of 
    their quasi-steady states and into dissipation times.
    This plot is to be considered in conjunction with
    Fig. \ref{lifetimeplot}. 
    \label{shockplot}}
  % Large surface density gradients become prominant
  % around vortex during dissipation} 
\end{figure}

@@

We characterize the smoothness of the outer gap edge with a dimensionless
gap edge gradient parameter
\begin{align}
  \delta\Sigma(t)= \la \frac{\partial \langle \Sigma(t,r)
    \rangle_{\phi}}{\partial r} \cdot \frac{r}{\langle \Sigma(t=0,r)
    \rangle_{\phi}} \ra_{\Delta r}  
\end{align}
where $\Delta r = r\in [r_p,r_p + 6r_h]$ is the radial range of averaging, spanning
from centre of the gap to the radius of the surface density maximum.
A larger $\delta \Sigma$ characterizes a sharper gap edge and  
larger local surface density maxima.

@@

This can be interpreted as the vortex providing a viscosity;  
and we measure a typical alpha viscosity $\alpha = O(10^{-2})$
associated with the vortex. This acts against gap-opening
by the planet, and smooths out the outer surface density bump,    
 so the condition for the RWI 
becomes less favourable. In
  order to re-launch the RWI, the surface density bump should
  reform. However, this is difficult as there is no more material
  in the planet's vicinity to clear out (Fig. \ref{gap_smoothed}) to
  form a surface density bump outside the gap. This may explain
why vortices do  not reform again (at least within the simulation
timescale).  After full decay the aspect ratio at the outer gap edge is
  $h\sim0.05$.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/gap_smoothness}
  \caption{Non-dimensional measure of the surface density gradient at
    the outer gap edge $\tilde\beta=1.0$. During the vortex quasi-steady state
    the gap edge is found to have a large gradient and sharp peak while vortex
    dissipation,
    which occurs at $t\approx10^3P_0$ as seen in Fig. \ref{lifetimeplot},
    works to smooth out the gap edge.
% A large $\delta\Sigma$
 %   corresponds to a more sharply peaked outer gap edge.
    \label{smoothnessplot}}  
\end{figure}

@@

We now examine vortex lifetimes as function of the imposed cooling
times. For this study, additional simulations with 
  $\tilde\beta=0.01,\,0.25,\,0.75,\,2.5,\,7.5$ were also performed.  
We define the vortex lifetime, $t_{\mathrm{life}}$, as the time
at which the over-density of the vortex
returns to $\Delta \Sigma/\Sigma_0\sim1$ \emph{after} reaching maximum
 (which is on the order of the
initial over-density associated with the gap formation).

@@

For fast cooling rates ($\tilde{\beta}\lesssim 1$), the
vortex lifetime is maximized for $\tilde{\beta}\to0$: we find 
$t_{\mathrm{life}} \approx 2100P_0$ for $\tilde{\beta}=0.01$ and 
decreases to $t_{\mathrm{life}} \approx 950P_0$ for
$\tilde{\beta}=0.25$. {\bf Note that for very small $\tilde{\beta}$, 
  there is significant contribution to the overall vortex lifetime due
  to a long decay timescale. }
For longer cooling times ($\tilde{\beta}\gtrsim
1$) the vortex lifetimes maximizes at $\tilde\beta=2.5$ with 
$t_{\mathrm{life}}\approx 1650P_0$.  
%The other timescales $t_\mathrm{diss}$ and
%$t_\mathrm{Mach}$ are less complicated, but still display a
%non-monotonic dependence on $\tilde{\beta}$, with a single maximum at
%$\tilde{\beta}=1$.

@@

\begin{figure}
  \includegraphics[width=\linewidth]{figures/betaplot}
  \caption{Characteristic timescales associated with vortex evolution, as a function of the cooling
    parameter $\tilde\beta$: time in which the $m=1$
    surface density amplitude begins to decay, $t_\mathrm{diss}$ (black);
    time at which the over-density at the
    vortex centre decreases to $\Delta\Sigma/\Sigma \sim 1$ after
    reaching maximum, $t_\mathrm{life}$ (red);
    and $t_\mathrm{Mach}$ is the time taken
    for the average Mach number around the vortex to maximise
    (blue). \label{betaplot}}  
\end{figure}

@@

%This effect can be seen in Mach number around the vortices where the
%growth of Mach number with respect to time is increasing with cooling rate.
%Thus the initially weaker formed vortices for higher cooling rate grow
%substantialy faster than the sound speed and can shock the system in earlier
%times.
%While our `planet-off' simulations show 
%This may be attributed to the higher temperature
%reached in the latter case 
%
%($h\simeq 0.07$) in comparison with the former
%($h\simeq 0.06$), since \cite{li00} showed that the RWI is favoured by
%larger disc temperature.

>>>>

\subsection{Gap structure}
We first compare the gap structures formed by planet-disc
interaction as a function of the cooling time. The azimuthally-averaged 
gap profiles are shown in Fig. \ref{intial1D} for different values of 
$\tilde\beta$. Gaps formed with lower $\tilde\beta$ (faster cooling) 
are deeper with steeper gradients at the gap edges. Faster cooling rates also 
increase (decrease) the surface density maxima (minima). However, a
clean gap does not form in this short time period.%
%\footnote{We find the
% RWI begins to develop if the disc is evolved beyond $t\sim30P_0$}.

@@

%However, notice the gap structure
%changes most significantly for $\beta =0.1 \to 1.0$. For $\beta \geq
%1$

@@

However, we remark that the change in the gap structure becomes less 
significant at long cooling times, as Fig. \ref{intial1D} shows that
the profiles with $\tilde{\beta}=1$ and $\tilde{\beta}=10$ are 
similar. This implies that the effect of cooling timescale on the
RWI through the set up of the gap profile, becomes less important for
large $\tilde{\beta}$. 
% or 'suggets that .... becomes unimportant for sufficiently large...'

@@

\subsection{Axisymmetric stability}
The initial planet-disc interaction form bumps and grooves in the gap profiles
which can potentially be unstable due to axisymmetric instabilities. The
generalised local axisymmetric stability condition is the Solberg-Hoiland
criterion,  
\begin{align}
  \kappa^2+N^2 \geq 0 
\end{align}
where
\begin{align}
 N^2=\frac{1}{\Sigma} \frac{\partial P}{\partial r}
 \left(\frac{1}{\Sigma} \frac{\partial \Sigma}{\partial
     r}-\frac{1}{\gamma P} \frac{\partial P}{\partial r}  \right) 
\end{align}
is the square of the Brunt-V\"ais\"al\"a frequency. 
At the outer gap edge $r=r_p+2.5r_h$,  where the RWI is excited
(see below),{\bf  we find $\kappa^2 + N^2$ reaches local minimum with a value
$\sim 0.45 \, \Omega^2(r_p)$
%$\sim 2\times10^{-4}$ (code units)
 for all $\tilde\beta$. The
Brunt-V\"ais\"al\"a frequency at the outer gap edge is
$N\sim 0.1 \, \Omega(r_p)$} 
% $N\sim 10^{-5}$,
decreasing marginally with longer cooling rate. 
The Solberg-Hoiland criteria is similarly satisfied for the entire 2D disc throughout the 
simulations.
Thus for all values of $\tilde\beta$ the planet-induced gaps are
stable to axisymmertic instabilities.

% {\bf Brunt frequency the same for all cooling?:small decrease with beta,
%  not more than order of magnitude}

%  {\bf i assume this means the azimuthally averaged
%   values. do you ever get $N^2<0$ anywhere in the 2D disc?:
%  This was not for azimuthal average, below 0 inside gap but kappa+N always above 0 everywhere}

\subsection{Non-axisymmetric instability}\label{linear}
We now examine the evolution of the gap for $t>30P_0$, with the
planet potential switched off, but with an added surface density
perturbation. For all three cooling times $\tilde{\beta}=0.1,\, 1.0,\,
10$, we observe exponential growth of non-axisymmetric
structures. An example is shown in Fig. \ref{linearmodes} for 
$\tilde{\beta}=10$. We characterise these
modes with an azimuthal wavenumber $m$ and growth rate $\sigma(m)$ as defined by
Eq.~\ref{fouriertransform}---\ref{growth}. Mode amplitudes were
averaged over $r-r_p\in[2,5]r_h$. Table \ref{modetable}
lists the growth rates measured during 
linear growth for 5 values of $m$ centred around that with maximum
growth rate.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=1.2cm
  0cm 0cm 0cm]{figures/linear_stability}
  \caption{Evolution of azimuthal Fourier modes of disc surface
    density, non-dimensionalized by the initial axisymmetric component 
    $\Sigma_0(t=0)$, for the
    `planet-off' simulation with $\tilde{\beta}=10$. Colours correspond
    to different $m$ values. The $m=3$ component is the fastest growing
    mode during linear growth with a growth rate of 
    $\gamma=0.017\Omega(r_p)$.\label{linearmodes}}
  % {\bf is the $\Sigma_0$ at t=0?} }  
\end{figure}

@@

\subsection{Long term evolution} \label{nonlinearplanetoff} 
% {\bf typical rossby numbers for these `planet-off' vortices? do
%   stronger vortices (beta=0.1) dissipate faster?:very complicated
% maybe not worth discussing since just comparison to planet-off}

@@

{\bf This decay is probably due to numerical viscosity. During the
  slow decay the vortex elongates (weakens) while its radial width
  remains $O(H)$, so its surface density decreases. In addition, for 
  $\tilde{\beta}=0.1,\,0.5$ and $1.0$ we also observe the appearance of
  spiral waves associated with the vortex, which may contribute to
  its dissipation (see below).} 
% (true?) (For isothermal and long
% cases yes. Spiral waves appear in beta=0.1,0.5,1.0}) 
We will see in the
next section that this {\bf decay after linear growth} is very
different to when the planet potential is kept on.

%Interestingly, though, vortices arising from the stronger
%instabilities (shorter cooling times or lower $\tilde{\beta}$) decays
%faster.

%main point: no planet: only get growth during linear instability
%           with palnet: further growth after linear regime

@@

\section{Non-linear evolution of
  gap-edge vortices with finite cooling time}\label{nonlinear} 
We now examine long-term simulations of gap-edge vortices for
$\tilde{\beta}=0.1,0.5,1,5,10$.{\bf(Additional cases are presented in 
  \S\ref{lifetime_discuss} when examining vortex lifetimes as a
  function of $\tilde{\beta}$.)} The planet potential is kept on
throughout. We employ a grid with $(N_r,N_{\phi})=(512,1024)$ in order
for these 
simulations to be computationally feasible. We also use a larger
disc with $r_{\mathrm{out}}=45r_{\mathrm{in}}$ to minimise boundary
effects on vortex evolution, and apply open boundaries at
$r=r_\mathrm{in},\,r_\mathrm{out}$.

We comment that lower-resolution simulations with
$(N_r,N_{\phi})=(256,512)$ show similar behaviour and trends as the
high-resolution runs reported below. {\bf These simulations
  include a comparison between a run with $\tilde{\beta}=0.01$ with a 
  locally isothermal disc, which show similar behavior. In particular,
  we find vortex lifetimes have a non-monotonic dependence on
  $\tilde{\beta}$, with longest lifetimes as $\tilde{\beta}\to 0$ (or
  locally isothermal discs) and for $\tilde{\beta} = O(1)$}.

@@

In all cases the system remains in a quasi-steady state for
$\gtrsim800P_0$ with a single vortex circulating 
the outer gap edge at the local Keplerian  
frequency. 
Fig. \ref{Vortex2D} shows a typical plot of the relative 
surface density perturbation in this state. During this stage, the 
vortex intensifies. This is better shown in Fig. \ref{rossbyplot} as the
evolution of Rossby numbers measured at the vortex centres.  
%As the vortex forms it has an initial 
%Rossby number $Ro\approx-0.1$ for all $\tilde\beta$
%(implying anti-cyclonic 
%motion). 
The Rossby number increases in magnitude during quasi-steady state,
but the maximum $|Ro|$ is similar for all $\tilde{\beta}$: the vortex
reaches a characteristic value of $Ro\approx-0.35$ for 
$\tilde\beta=0.1$ and $Ro\approx -0.45$ for $\tilde\beta=10$.

@@

\begin{figure}
  \includegraphics[width=\linewidth,height=\linewidth]{figures/vortex2D}
  \caption{Relative surface density perturbation for the
    $\tilde\beta=0.1$ case during quasi-steady state with a single
    vortex at the outer gap edge. The plot for other values of the cooling time
    $\tilde{\beta}$ are similar. {\bf The decrease in the surface
      density near} $r\sim 40 r_\mathrm{in}$ {\bf arises from mass loss due to the open
      boundary condition imposed at} $r_\mathrm{out}$.
    \label{Vortex2D} }
\end{figure}

@@

% We remark that the increase in the
% vortex surface density directly due to the pile-up of material at the
% gap edge (because of gap-opening) is small: the average surface density
% perturbation at vortex radial region, $r\sim r_{p}+5r_h$, is 
% $\langle\Delta\Sigma/\Sigma_0\rangle_\phi<1$.

@@

% {\bf may be more appropriate to quote the azimuthally averaged
%   perturbation at the location of the vortex (5-6 Hill radii from the
%   planet?): yes, this was my original idea, mistyped as 2.5
% }

% {\bf any trend between maximum over-density achieved and cooling?: generally increasing with beta, max at beta=5}

% {\bf question: vortex amplitude plot suggest just after linear growth,
%   vortex amplitudes are smaller with increasing beta, but later on in
%   the quasi-steady state, this trend reverses? yes. 
% }

% {\bf Perhaps a 2D snapshot of what the gap edge looks like after
%   vortex death (like the one in the poster).: Not sure if this is figure 
% you were thinking (Fig 8)
% % defer definition of  vortex lifetime to later, when we discuss it. 
% }

@@

In quasi-steady ($t=700P_0$) the vortex is elongated with a vortex aspect ratio $ \approx 4$
, but becomes more compact approaching a ratio of 2 during its decay ($t=1310P_0$).   
Notice in Fig. \ref{shockplot} the appearance of wakes extending from
either side of the vortex at $t=1300P_0$. These 
wakes correlate with large gradients in surface density (bottom
panel), and are first seen in the later half of the quasi-steady
state. We find the time at which 
the vortex begins to decay coincides with the emergence of these
wakes. %{\bf true statement? yes, distortion of vortex shape form wakes happens at same time}

{\bf During the quasi-steady state the vortex 
  orbits at $r\sim r_p+6r_h$. We do not see significant
  vortex migration at this stage, since the vortex is located at a 
  surface density maximum \citep{paardekooper10}. However,
  simultaneous with the appearance of the wakes, we observe the vortex
  begins to migrate inwards to $r\sim r_p+5r_h$.  
  % This is a result of the the spiral waves transporting
  % angular momentum to the disc (Is this true? Shocks on the inner disc
  % seem stronger so I would think it would migrate outward). 
  }

% As the vortices grow in intensity wakes are seen to extend from them. 
During quasi-steady state the average value of the surface density gradient
along the wakes is $|w_s\nabla\Sigma/\Sigma| \sim 0.4 $, where 
$w_s\simeq0.1$ (code units) is a typical length scale of 
the surface density variation across the wake.   
Just before the $m=1$ amplitude begins to decrease, we observe this quantity
sharply increases to $ \sim 0.6 $, and remains around this value
until the vortex dies out, at which point the associated Rossby number
begins decreasing to zero. {\bf After the 
  vortex reaches small amplitudes ($1\lesssim\Delta\Sigma/\Sigma$), it 
  migrates out to $r\sim r_p+ 6.5 r_h$. 
}
%{\bf :check}.

@@

\begin{figure}
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/shock1}
  }
  \hfill
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/shock2}
  }
  \hfill
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/shock3}
  } \\[-0.98cm]
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/shock4}
  }
  \hfill
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/shock5}
  }
  \hfill
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/shock6}
  } 
  \caption{The vortex in the case with $\tilde{\beta}=1$
    during quasi-steady state (left), start of decay 
    (middle), and just after the decay in the $m=1$ amplitude
    (right). The surface density perturbation
    (top) and the associated surface density gradient (bottom) are
    shown. Wake like features corresponding to large density gradients
    are found to originate from the vortices during the late phase of 
    their quasi-steady states and into dissipation times.
    This plot is to be considered in conjunction with
    Fig. \ref{lifetimeplot}. 
    \label{shockplot}}
  % Large surface density gradients become prominant
  % around vortex during dissipation} 
\end{figure}

@@

We characterise the smoothness of the outer gap edge with a dimensionless
gap edge gradient parameter
\begin{align}
  \delta\Sigma(t)= \la \frac{\partial \langle \Sigma(t,r)
    \rangle_{\phi}}{\partial r} \cdot \frac{r}{\langle \Sigma(t=0,r)
    \rangle_{\phi}} \ra_{\Delta r}  
\end{align}
where $\Delta r = r\in [r_p,r_p + 6r_h]$ is the radial range of averaging, spanning
from centre of the gap to the radius of the surface density maximum.
A larger $\delta \Sigma$ characterizes a sharper gap edge and  
larger local surface density maxima.

@@

This can be interpreted as the vortex providing a viscosity;  
and we measure a typical alpha viscosity $\alpha = O(10^{-2})$
associated with the vortex. This acts against gap-opening
by the planet{\bf, and smooths out the outer surface density bump,    
} so the condition for the RWI 
becomes less favourable. {\bf In
  order to re-launch the RWI, the surface density bump should
  reform. However, this is difficult as there is no more material
  in the planet's vicinity to clear out (Fig. \ref{gap_smoothed}) to
  form a surface density bump outside the gap.} {\bf This may} explain
why vortices do  not reform again (at least within the simulation timescale). 
{\bf After full decay the aspect ratio at the outer gap edge is
  $h\sim0.05$. }
% (0.06 for beta=5.0,10)}

% what is aspect-ratio at outer gap edge after full decay?

%Thus, the
%time needed for the vortex to grow to sufficient amplitude to modify
%the gap structure

\begin{figure}
  \includegraphics[width=\linewidth]{figures/gap_smoothness}
  \caption{Non-dimensional measure of the surface density gradient at
    the outer gap edge $\tilde\beta=1.0$. During the vortex quasi-steady state
    the gap edge is found to have a large gradient and sharp peak while vortex
    dissipation,
    which occurs at $t\approx1000P_0$ as seen in Fig. \ref{lifetimeplot},
    works to smooth out the gap edge.
% A large $\delta\Sigma$
 %   corresponds to a more sharply peaked outer gap edge.
    \label{smoothnessplot}}  
\end{figure}

@@

We now examine vortex lifetimes as function of the imposed cooling
times. For this study, additional simulations with 
  $\tilde\beta=0.01,\,0.25,\,0.75,\,2.5,\,7.5$ were also performed.  
We define the vortex lifetime, $t_{\mathrm{life}}$, as the time
at which the over-density of the vortex
% to reach $10\%$ of its max overdensity value
returns to $\Delta \Sigma/\Sigma_0\sim1$ \emph{after} reaching maximum
%{\bf correct interpretation?}
 (which is on the order of the
initial over-density associated with the gap formation).

@@

For fast cooling rates ($\tilde{\beta}\lesssim 1$), the
vortex lifetime is maximized for $\tilde{\beta}\to0$: we find 
$t_{\mathrm{life}} \approx 2100P_0$ for $\tilde{\beta}=0.01$ and 
decreases to $t_{\mathrm{life}} \approx 950P_0$ for
$\tilde{\beta}=0.25$. {\bf Note that for very small $\tilde{\beta}$, 
  there is significant contribution to the overall vortex lifetime due
  to a long decay timescale. }
For longer cooling times ($\tilde{\beta}\gtrsim
1$) the vortex lifetimes maximizes at $\tilde\beta=2.5$ with 
$t_{\mathrm{life}}\approx 1650P_0$.

@@

\begin{figure}
  \includegraphics[width=\linewidth]{figures/betaplot}
  \caption{Characteristic timescales associated with vortex evolution, as a function of the cooling
    parameter $\tilde\beta$: time in which the $m=1$
    surface density amplitude begins to decay, $t_\mathrm{diss}$ (black);
    time at which the over-density at the
    vortex centre decreases to $\Delta\Sigma/\Sigma \sim 1$ after
    reaching maximum, $t_\mathrm{life}$ (red);
    and $t_\mathrm{Mach}$ is the time taken
    for the average Mach number around the vortex to maximise
    (blue). \label{betaplot}}  
% {\bf maybe useful to over-plot other timescales,
%      e.g. time taken to reach max mach, time for rossby number to
%      reach zero.}
\end{figure}

@@

%This effect can be seen in Mach number around the vortices where the
%growth of Mach number with respect to time is increasing with cooling rate.
%Thus the intially weaker formed vortices for higher cooling rate grow
%substantialy faster than the sound speed and can shock the system in earlier
%times.
%While our `planet-off' simulations show 
%This may be attributed to the higher temperature
%reached in the latter case 
%
%($h\simeq 0.07$) in comparison with the former
%($h\simeq 0.06$), since \cite{li00} showed that the RWI is favoured by
%larger disc temperature.
